A Middleware for Symmetric Peer-to-Peer Applications

### 2023-03-12 tpds-23/tpds-23.tex
...

\IEEEtitleabstractindextext{%
\begin{abstract}
In real-time networked applications, such as collaborative documents, users can
interact remotely and yet share the exact same experience as if they were using
a single machine.
%
In this work, we propose a middleware for these \emph{symmetric peer-to-peer
applications} in which decentralized instances can broadcast asynchronous
events and yet conform to identical behavior.
%
Peers are allowed to join and leave the network at any time.
Application developers must adhere to a deterministic and stateless API.
%, and only supports pre-allocated memory.
%
The middleware is responsible for synchronizing the events in a global shared
timeline across the network.
Also, based on memory snapshots and deterministic simulation, a ``time machine''
can rollback conflicting peers to resynchronize their states.
%
We show that the middleware can handle applications with over 20 peers in
heterogeneous topologies under high churns.
For instance, in a scenario with a network latency of 50ms and 25 events per
minute, the peers need to roll back only 3\% of the time.
\end{abstract}

\begin{IEEEkeywords}
Bitcoin, blockchains, CRDT, distributed consensus, peer-to-peer,
publish-subscribe, reputation system, VCS
\end{IEEEkeywords}}

@@

Real-time networked applications allow multiple users to interact remotely and
yet share the same experience.
Examples of these \emph{symmetric distributed applications}~\cite{gals} are
collaborative documents, watch parties, and multi-player games.
%~\cite{TODO:collab}.
%
In order to reproduce the exact behavior in multiple devices, the application
must be able to synchronize time and execution across the network.
In addition, since users can interact asynchronously with the application,
event occurrences must somehow be synchronized back across devices.

A common approach towards symmetric applications is to introduce a middleware
to orchestrate events and time in the network~\cite{gals,croquet}.
This way, applications developers can rely on middleware primitives to trigger
events, which are intercepted and synchronized in a global shared timeline
across the network.
Developers must also restrict themselves to deterministic and stateless calls
only, such that execution can be equally reproduced in all nodes according to
the shared timeline. %~\cite{TODO:det_reprod}.
However, current solutions depend on a central server to interconnect network
nodes and determine a shared timeline.
Therefore, applications may experience failures, unless users are permanently
connected, with no partitions in the network, particularly from the server.
In contrast, a decentralized solution can be resilient to partitions, and
ultimately, even work completely offline, as required by local-first
software~\cite{local}.

In this work, we propose a middleware for \emph{symmetric peer-to-peer
applications}, which do not rely on central servers for coordination.
Peers in the application form a dynamic network graph and communicate only
with direct neighbours, as in typical unstructured peer-to-peer
networks~\cite{p2p.survey}.
Events are flooded in the graph and are applied with a small deadline delay to
compensate the network latency.
To deal with events received out of order or too late, a distributed time
machine can rollback peers to previous states and reapply the events in order,
such that a single consistent timeline is established across the network.
Our main contribution is the design of a middleware to ensure that all peers
    (a) meet event deadlines,
    (b) advance in sync in real time, and
    (c) can leave and join the network and remain symmetric.

@@

In this section, we revisit existing solutions for symmetric distributed
applications, which are not necessarily peer to peer.
We focus on
    (a) how the network is organized,
    (b) how global time is determined, and
    (c) how events are propagated and applied.
Figure~\ref{fig.table} compares three selected works regarding these aspects.
%
At the end of this section, we also discuss related work on local and
distributed time machines.

@@

Croquet~\cite{croquet,croquet.site} guarantees bit-identical real-time behavior
for users in collaborative distributed environments.
%
As detailed in Figure~\ref{fig.algos}, a centralized \emph{reflector} issues
periodic ticks, such that all nodes in the network advance together according
to a synchronized global shared clock.
If a user generates an event, it is sent back to the reflector, which
broadcasts the event in the next tick, which all nodes apply in sync.
%
Croquet takes periodic snapshots of the whole application state in order to
support late joins to a running session.
The new node just needs to request the latest snapshot and simulate the
remaining events to reach the current running state.
%
As indicated in Figure~\ref{fig.table}, Croquet relies on a centralized
network (a), in which nodes advance in sync (b), and in which event broadcasts
and outcomes depend on the central server to be online (c).

@@

In both solutions, the network advances together in real time as a whole, with
a total order among events, which is determined by a central server that must
be permanently online.
All clients compute events in sequence, respecting timestamps, and using only
deterministic and stateless calls.
This way, it is guaranteed that all clients reach the same state, and
observe bit-identical steps.

In this work, as indicated in Figure~\ref{fig.table}, our goal is to support
peer-to-peer (a) real-time symmetric execution, while still tolerating
out-of-order (b) and offline (c) event generation.
%
As detailed in Figure~\ref{fig.algos}, peers have independent timelines and
can communicate events directly to each other.
Event deadlines are delayed to be able to reach other peers in time.
In the case a deadline is missed, the peer rolls back in time and then fast
forwards execution until it catches real-time behavior again.
%
Like Croquet and GALS, peers compute events in sequence using deterministic and
stateless calls only.

@@

An antagonistic approach to deterministic reactions to events is to model the
applications with Conflict-free Replicated Data Types~\cite{crdts}.
CRDTs are designed in such a way that all operations are commutative, so that
the order in which they are applied does not affect the final outcome.
%
As detailed in Figure~\ref{fig.algos}, peers can communicate operations
directly to each other with no central authority.
Also, since operations need not to be ordered, they can be applied at the
very first moment they are generated or received, even if the originating peers
are offline.
%
Note that it is not possible to timestamp commutative operations under a unique
global shared clock.
Hence, there is no notion of a timeline in which peers go through bit-identical
steps, but they do eventually reach the same state~\cite{crdts.eventual}.
%
On the one hand, the main advantage of CRDTs is to support local-first
software~\cite{local}, since they can work offline in the same way as online.
On the other hand, they provide very restrictive APIs (the CRDTs themselves),
and do not support real-time consensus among peers~\cite{crdts.consensus}.
%
As indicated in Figure~\ref{fig.table}, CRDTs work on decentralized networks
(a), in which peers advance independently (b), and in which operations can be
applied immediately, even while offline (c).
%
Automerge along with its accompanying Hypermerge peer-to-peer infrastructure
is an example of decentralized CRDTs~\cite{p2p.automerge,p2p.pushpin}.

@@

Regarding decentralized time machines, Fusion~\cite{fusion} is a state
synchronization networking library for the Unity game engine.
It provides networked tick-based simulation in two modes: client-server
(\emph{hosted mode}) and peer-to-peer (\emph{shared mode}).
The hosted mode is based on continuous memory snapshots, which allows for
full rollbacks when clients diverge.
However, the shared mode is less powerful and only supports eventual
consistency among clients.
We presume that continuous memory snapshots would be too costly without a
central server.
%
Regarding single-user local time machines, the video game Braid%
\footnote{Braid: \url{http://braid-game.com/}}
is designed on the assumption that players have unrestricted rollbacks as part
of the game mechanics.
Like Fusion, the implementation is based on continuous memory snapshots,
instead of deterministic simulation as we propose in this work.
%
Schuster~\cite{tml.js} proposes a simple API for time travelling in
event-driven JavaScript:
    \code{init()} to return an initial state;
    \code{handle(evt,mem)} to process events and modify state; and
    \code{render(mem)} to output the current state.
As we describe next, we use a similar approach by separating state mutation
from rendering~\cite{tml.alive}.

@@

\begin{figure}
    \centering
    \includegraphics[width=\linewidth]{move}
    \caption[XXX] {
        A moving rectangle is controlled by peers connected indirectly.
        Users can press \code{↑ ↓ → ←} to change the direction of the
        rectangle or \code{SPACE} to pause the application.
        The inputs are applied simultaneously in all peers, as if they were
        mirrors of a single application.
        Video: \url{http://youtube.com/TODO-hidden-blind-review}
        \label{fig.move}
    }
\end{figure}

We guide our discussion through the didactic example of Figure~\ref{fig.move},
in which multiple users control the direction of a moving rectangle using the
arrow keys.
The six instances are connected in a peer-to-peer network, such that most
peers can only communicate indirectly to each other.
%
The application is symmetric in the sense that if any user presses a key, a
corresponding event is broadcast and applied in all peers simultaneously, as
if they were a single mirrored machine.
At any moment, if a user pauses the application, all peers are guaranteed to
pause exactly at the same frame, with the rectangle ``bit-identically'' at the
very same position.

\begin{figure}
  \centering
  \includegraphics[width=\linewidth]{middleware}
  \caption{
    Applications generate asynchronous events and communicate through an API
    with the middleware.
    The middleware orchestrates the peers, synchronizing their ticks locally
    and broadcasting events with a deadline delay ($\Delta$).
    \label{fig.middleware}
  }
\end{figure}

In Figure~\ref{fig.middleware}, a peer is depicted as an application using an
API to communicate with the middleware.
The application itself is not aware of the network and has no direct access to
other peers, which communicate transparently through the middleware.
The application generates asynchronous events that go through the middleware,
which synchronizes them with a timestamp with a deadline delay ($\Delta$)
scheduled to the future, such that all peers can satisfy.
The middleware controls the execution of the application by issuing ticks and
synchronized events.
Note that ticks need not to be broadcast, only $\Delta$ events.

The main job of the middleware is to deal with the uncertainties of the
network in order to provide a consistent timeline and preserve the symmetric
behavior across peers.
As described in the Introduction, the middleware ensures that all peers
    (a) meet event deadlines,
    (b) advance in sync in real time, and
    (c) can leave and join the network and remain symmetric:
%
\begin{enumerate}
\item \textbf{Event Deadlines:}
Due to the inherent latency of networks, peers generate events with a
deliberate $\Delta$ so that they can reach other peers in time to be applied in
sync.
%We first assume a predetermined delay for the sake of simplicity.
Nevertheless, a peer ahead of time may receive an event that should have been
applied in the past. %, even considering its $\Delta$.
In this case, the middleware automatically rolls back the peer, applies the
event at the correct time, and then fast forwards the peer, re-applying all
remaining events up to the real time.
As we detail further, rolling back a peer requires to simulate the whole
application execution and event occurrences from the beginning.
As an optimization, the middleware takes periodic snapshots locally to avoid
full simulation.
%
\item \textbf{Time Synchronization:}
Since peers run in different machines, their timelines will inevitably be out
of sync because applications are launched locally at different times, and
because internal clocks may diverge over time (e.g., from rollbacks or timer
polling inaccuracies).
The middleware assumes the maximum time among all peers the be the
``correct real time'', so that other peers must fast forward to catch up with
it.
Therefore, in order to determine the maximum time and synchronize the clocks,
the middleware proceeds as follows:
    (a) peers broadcast \code{SYNC} events periodically with their local
        times, and
    (b) if a peer is behind a received \code{SYNC}, it advances its frames
        proportionally faster to catch up in time.
%
To avoid abrupt visual changes, the middleware makes smooth time transitions
that take up to $1s$ to complete.
%
\item \textbf{Peer Churn:}
Considering that we target dynamic peer-to-peer networks, it is important that
the middleware can handle arrival and departure of peers seamlessly.
Note that high churn may lead to intermittent network partitions.
Given our unstructured approach, nothing needs to be done when a peer leaves
the network.
However, when a peer joins the network, the middleware needs to ensure that it
receives from other peer all events that ever happened, in order.
This is trivial since all peers already need to hold the full event history (as
described in item (a)).
% including their absolute timestamps in the shared timeline.
\end{enumerate}

@@

Note that using callbacks is a common technique for event-driven
programming in Games, GUIs, and interactive applications in general, sharing
similarities with \emph{MVC}, \emph{Publish-Subscribe}, and \emph{Observer}
patterns~\cite{meyer,nystrom}.
%
Furthermore, the default run-to-completion semantics of callbacks is key to
ensure the desired determinism, since unlike threads, they run sequentially and
atomically from start to end~\cite{events,threads}.

@@

If dynamic allocation is required, it is necessary to hold a finite memory
pool in \code{G} with a custom allocator.
This is a limitation of C, since languages with serialization and
meta-programming offer better mechanisms to take memory snapshots.
Another alternative is to provide a serialization callback, which we left for
future work.

@@

The middleware calls \code{cb\_ini} twice: at the beginning and at the end of
the loop.
The callback should initialize (and finalize) the network topology, as well as
static immutable globals that can live outside the simulation memory, such as
the window and image textures in our example:

@@

@@

Time travelling is also the reason why \code{cb\_sim} must be separated from
the rendering callback \code{cb\_ren}~\cite{tml.js}, which would otherwise
render the screen multiple times while travelling.
The middleware renders the screen after each frame in real time.
%, which the middleware calls only on real-time frames.
In our example, \code{cb\_ren} just needs to redraw the rectangle at the
current position:

@@

The callback \code{cb\_evt} is called by the middleware in real time, whenever
a local SDL event occurs.
This callback has two goals:
    (a) map and broadcast application events, and
    (b) provide instant feedback to the user.
%As mentioned previously,
Not all local low-level events need to broadcast application events in the
network.
The callback is allowed to perform side effects, such as modifying the network
topology, or exhibiting instant feedback on the screen.
In our example, we only generate application events for the 5 key events of
interest, but we also exhibit the pressed key in real time as a visual
feedback:

@@

We now detail how the middleware orchestrates an application and ensures that
all peers
    (a) meet event deadlines,
    (b) advance in sync in real time, and
    (c) can leave and join the network and remain symmetric.

@@

As illustrated in Figure~\ref{fig.middleware}, events are broadcast between
peers with a deadline $\Delta$ scheduled to the future, such that all peers are
able to apply them in sync.
%
To prevent broadcast cycles, each peer keeps a collection of the highest
timestamps it received from each other peer, such that lower numbers can be
ignored when received.
Since we rely on TCP connections, it is not possible to receive out-of-order
timestamps attached to a given source peer.
When the middleware receives a higher number from a given peer, it updates this
number and triggers a broadcast to neighbours.
%
%As mentioned in Section~\ref{sec.tml.api}, we assume that peers assign
%themselves unique identifiers in a contiguous range, such that they fit in a
%simple vector of timestamps.
%Therefore, this is not a severe limitation, and a separate mechanism could be
%used for assignments, with few modifications to the core of the middleware.

@@

The middleware main loop \code{p2p\_loop} is responsible for calling the
application callbacks, and also controlling its timeline.
The loop continuously checks for network packets, local inputs, and time ticks,
as follows:

@@

The fact that peers share a global timeline of event, and that they go through
bit-identical steps ensures consistency in the network.
However, since the clocks may diverge temporarily, we must rely on rollbacks.
Therefore, the key to preserve consistency in the presence of rollbacks is to
restrict our programming API to deterministic and stateless calls only.
This way, it becomes impossible to diverge from other peers by, for example,
re-executing stateful file operations multiple times (or at different times).

@@

We run the application main loop at 50 FPS ($20ms$ per frame) and calculate
the middleware overhead at each iteration.
We found a negligible overhead of $9us$ in the average, which corresponds to
less than $1/2000$ of the CPU time.
%
In order to have an absolute clock, we use a single machine to simulate all
peers, allowing us to merge their logs into a single file with a comparable
shared timeline.
%
We use \emph{NetEm}~\cite{netem} to simulate the network latency with a normal
distribution.
% (e.g., \code{tc qdisc add dev lo root netem delay 200ms 40ms distribution normal}).
%
We vary
    (i)   the network latency ($5-500ms$ between each peer),
    (ii)  the rate of events ($5-200~evt/sec$), and
    (iii) the deadline $\Delta$ multiplier ($1-5x$).
%
Figure~\ref{fig.mult} shows the $\Delta$s considering the network latency and
chosen multiplier.
For instance, a $50ms$ latency with a $2x$ multiplier results in
$\Delta=500ms$, i.e., each event is triggered with this deadline delay to
compensate the given latency and fixed average of 5 hops
($5 \times 50 \times 2 = 500$).

>>>>

\IEEEtitleabstractindextext{%
\begin{abstract}
In real-time networked applications, such as collaborative documents, users can
interact remotely and yet share the exact same experience as if they were using
a single machine.
%
In this work, we propose a middleware for \emph{symmetric peer-to-peer
applications} in which decentralized instances can broadcast asynchronous
events and yet conform to identical behavior.
%
Peers are allowed to join and leave the network at any time.
Application developers must adhere to a deterministic and stateless API.
%, and only supports pre-allocated memory.
%
The middleware is responsible for synchronizing the events in a global shared
timeline across the network.
Also, based on memory snapshots and deterministic simulation, a ``time machine''
can rollback conflicting peers to resynchronize their states.
%
We show that the middleware can handle applications with over 20 peers in
heterogeneous topologies under high churns.
For instance, in a scenario with a network latency of 50ms and 25 events per
minute, the peers need to roll back only 3\% of the time to remain fully
symmetric.
\end{abstract}

\begin{IEEEkeywords}
collaboration, determinism, peer-to-peer, time machine
\end{IEEEkeywords}}

@@

Real-time networked applications allow multiple users to interact remotely and
yet share the same experience.
Examples of these \emph{symmetric distributed applications}~\cite{gals} are
collaborative documents, watch parties, and multi-player games.
%~\cite{TODO:collab}.
%
In order to reproduce the exact behavior in multiple devices, the application
must be able to synchronize time and execution across the network.
In addition, since users can interact asynchronously with the application,
event occurrences must somehow be synchronized back across devices.
%
This problem relates to consistency in distributed systems, and is also known
as \emph{state melding}~\cite{melding} in the context of virtual worlds, which
adds the requirement of a time deadline to reach consistency.

A common approach towards symmetric applications is to introduce a middleware
to orchestrate events and time in the network~\cite{gals,croquet}.
This way, applications developers can rely on middleware primitives to trigger
events, which are intercepted and synchronized in a global shared timeline
across the network.
Developers must also restrict themselves to deterministic and stateless calls
only, such that execution can be equally reproduced in all nodes according to
the shared timeline. %~\cite{TODO:det_reprod}.
However, current solutions depend on a central server to interconnect network
nodes and determine the shared timeline.
Therefore, applications may experience failures, unless users are permanently
connected, with no partitions in the network, particularly from the server.
In contrast, a decentralized solution could be resilient to partitions, and
ultimately, even work offline as promoted by local-first software~\cite{local}.

In this work, we propose a middleware for \emph{symmetric peer-to-peer
applications}, which do not rely on central servers for coordination.
Peers in the application form a dynamic network graph and communicate only
with direct neighbours, as in typical unstructured peer-to-peer
networks~\cite{p2p.survey}.
%
Events are flooded in the graph and are applied with a small deadline delay to
compensate the network latency.
To deal with events received out of order or too late, a time machine can
rollback a peer to a previous state and reapply the events in order, such that
a single consistent timeline is established across the network.
%
Our main contribution is the design of a middleware to ensure that all peers
    (a) meet event deadlines,
    (b) advance in sync in real time, and
    (c) can leave and join the network and remain symmetric.
%
Our middleware provides \emph{perceptive consistency}~\cite{melding}, which
ensures that an event triggered by any peer at time $t$ is perceived by all
peers simultaneously at time $t+\Delta$.
%
As the main limitation, the middleware cannot ensure \emph{instantaneity} for
events in most scenarios, which would require $\Delta$ to be less than a
perceptive threshold~\cite{melding}.
We also that assume peers are non malicious in the sense that they do not lie
about the time in which events occurred locally.

@@

In this section, we revisit existing solutions for symmetric distributed
applications, which are not necessarily peer to peer.
We focus on
    (a) how the network is organized,
    (b) how global time is determined, and
    (c) how events are propagated and applied.
Figure~\ref{fig.table} compares three selected works regarding these aspects
against our work.
%
One of these approaches relies on CRDTs~\cite{crdts}, and is discussed in
separate.
%
At the end of this section, we also discuss related work on local and
distributed time machines.

@@

Croquet~\cite{croquet,croquet.site} guarantees bit-identical real-time behavior
for users in collaborative distributed environments.
%
As detailed in Figure~\ref{fig.algos}, a centralized \emph{reflector} issues
fixed periodic ticks (e.g., every $50ms$), such that all nodes in the network
advance together according to a synchronized global shared clock.
If a user generates an event, it is sent back to the reflector, which
broadcasts the event in the next tick, which all nodes apply in sync.
%
This mechanism is similar to \emph{bucket synchronization}~\cite{gautier}, in
which events are assigned to time buckets to be applied later.
%
Croquet takes periodic snapshots of the whole application state in order to
support late joins to a running session.
The new node just needs to request the latest snapshot and simulate the
remaining events to reach the current running state.
%
As indicated in Figure~\ref{fig.table}, Croquet relies on a centralized
network (a), in which nodes advance in sync (b), and in which event broadcasts
and outcomes depend on the central server to be online (c).

@@

In both solutions, the network advances together in real time as a whole, with
a total order among events, which is determined by a central server that must
be permanently online.
All clients compute events in sequence, respecting timestamps, and using only
deterministic and stateless calls.
This way, it is guaranteed that all clients reach the exact same state, and
observe bit-identical steps.

In this work, as indicated in Figure~\ref{fig.table}, our goal is to support
    (a) peer-to-peer real-time symmetric execution, while still
    (b) tolerating out-of-order, and
    (c) offline event generation.
%
As detailed in Figure~\ref{fig.algos}, peers have independent timelines and
can communicate events directly to each other.
Event deadlines are delayed to be able to reach other peers in time.
In the case a deadline is missed, the peer rolls back in time and then fast
forwards execution until it catches real-time behavior again.
%
This technique is also known as \emph{log-based
rollback-recovery}~\cite{rollback}, but applied in the context of real-time
collaborative applications.
%
Like Croquet and GALS, peers compute events in sequence using deterministic and
stateless calls only.

As it is the case for the discussed work, we consider that middlewares for
symmetric applications, which provide bit-identical behavior, must comply with
\emph{perceptive consistency}~\cite{melding}, as follows:

``\textbf{Perceptive consistency}~\cite{qin,bouillot}.
If any operation is executed on process $i$ at time $t$, it must be executed on
process $j$ at $t$ as well.
At a high-level point of view, it indicates that for a group of users
interacting with each other, if a user takes an action, then the action will be
perceived by everyone simultaneously (though the action might be perceived at a
time later than the time at which it was taken by the user). It is also
referred to as absolute consistency~\cite{qin}.''

@@

An antagonistic approach to deterministic reactions to events is to model the
applications with Conflict-free Replicated Data Types~\cite{crdts}.
CRDTs are designed in such a way that all operations are commutative, so that
the order in which they are applied does not affect the final outcome.
%
As detailed in Figure~\ref{fig.algos}, peers can communicate operations
directly to each other with no central authority.
Also, since operations need not to be ordered, they can be applied at the
very first moment they are generated or received, even if the originating peers
are offline.
%
Note that it is not possible to timestamp commutative operations under a unique
global shared clock.
Hence, there is no notion of a timeline in which peers go through bit-identical
steps, but they do eventually reach the same state~\cite{crdts.eventual}.
%
On the one hand, the main advantage of CRDTs is to support local-first
software~\cite{local}, since they can work offline in the same way as online.
On the other hand, they provide very restrictive APIs (the CRDTs themselves),
and do not support real-time consensus among peers~\cite{crdts.consensus}.
%
As indicated in Figure~\ref{fig.table}, CRDTs work on
    (a) decentralized networks, in which
    (b) peers advance independently, and in which
    (c) operations can be applied immediately, even while offline.
%
Automerge~\cite{p2p.automerge} along with its accompanying
Hypermerge~\cite{p2p.pushpin} peer-to-peer infrastructure is an example of
decentralized CRDTs.

CRDTs provide \emph{delayed consistency}~\cite{melding}, as follows:

``\textbf{Delayed consistency}~\cite{qin}.
This is a relaxation of perceptive consistency.
In this model, for any operation issued by host $i$, say at time $t$, it takes
effect instantaneously on host $i$.
On a remote host $j$, it is executed at time $t + (i,j)$, where $(i,j)$ is the
phase different between processes $i$ and $j$, and it is a parameter specific
to these two processes.
Unlike perceptive consistency, this model does not force all processes to
execute the same operation at the same time.
Rather, it makes sure that the state of objects operated by one process be
viewed at another process with a consistent delay.''

@@

Regarding decentralized time machines, the \emph{Time Warp}
mechanism~\cite{timewarp} rollbacks the memory state of a replica when it
receives an event from the past, and also broadcasts ``anti-messages'' for all
events that were incorrectly applied.
Our work assumes that all events are correct, and reapplies them relying on
deterministic simulation.

Fusion~\cite{fusion} is a state synchronization networking library for the
Unity game engine.
It provides networked tick-based simulation in two modes: client-server
(\emph{hosted mode}) and peer-to-peer (\emph{shared mode}).
The hosted mode is based on continuous memory snapshots, which allows for
full rollbacks when clients diverge.
However, the shared mode is less powerful and only supports eventual
consistency among clients.
We presume that continuous memory snapshots would be too costly without a
central server.

Regarding single-user local time machines, the video game Braid%
\footnote{Braid: \url{http://braid-game.com/}}
is designed on the assumption that players have unrestricted rollbacks as part
of the game mechanics.
Like Fusion, the implementation is based on continuous memory snapshots,
instead of deterministic simulation as we propose in this work.
%
Schuster~\cite{tml.js} proposes a simple API for time travelling in
event-driven JavaScript:
    \code{init()} to return an initial state;
    \code{handle(evt,mem)} to process events and modify state; and
    \code{render(mem)} to output the current state.
As we describe next, we use a similar approach by separating state mutation
from rendering~\cite{tml.alive}.

@@

\begin{figure}
    \centering
    \includegraphics[width=\linewidth]{move}
    \caption[XXX] {
        \textbf{An application example:}
        A moving rectangle is controlled by peers connected indirectly.
        Users can press \code{↑ ↓ → ←} to change the direction of the
        rectangle or \code{SPACE} to pause the application.
        The inputs are applied simultaneously in all peers, as if they were
        mirrors of a single application.
        Video: \url{http://youtube.com/TODO-hidden-blind-review}
        \label{fig.move}
    }
\end{figure}

We guide our discussion through the didactic example of Figure~\ref{fig.move},
in which multiple users control the direction of a moving rectangle using the
arrow keys.
The six instances are connected in a peer-to-peer network, such that most
peers can only communicate indirectly to each other.
%
The application is symmetric in the sense that if any user presses a key, a
corresponding event is broadcast and applied in all peers simultaneously, as
if they were a single mirrored machine.
At any moment, if a user pauses the application, all peers are guaranteed to
pause exactly at the same frame, with the rectangle ``bit-identically'' at the
very same position.
%
This is only possible because the middleware ensures that all peers shares the
same timeline of events, in which only deterministic operations are applied.

\begin{figure}
  \centering
  \includegraphics[width=\linewidth]{middleware}
  \caption{
    \textbf{The middleware architecture:}
    Applications generate asynchronous events and communicate through an API
    with the middleware.
    The middleware orchestrates the peers, synchronizing their ticks locally
    and broadcasting events with a deadline delay ($\Delta$).
    \label{fig.middleware}
  }
\end{figure}

In Figure~\ref{fig.middleware}, a peer is depicted as an application using an
API to communicate with the middleware.
The application itself is not aware of the network and has no direct access to
other peers, which communicate transparently through the middleware.
The application generates asynchronous events that go through the middleware,
which synchronizes them with a timestamp with a deadline delay ($\Delta$)
scheduled to the future, such that all peers can satisfy (also known as
\emph{local lag}~\cite{mauve}]).
The middleware controls the execution of the application by issuing ticks and
synchronized events.
Note that ticks need not to be broadcast, only $\Delta$ events.

The main job of the middleware is to deal with the uncertainties of the
network in order to provide a consistent timeline and preserve the symmetric
behavior across peers.
As described in the Introduction, the middleware ensures that all peers
    (a) meet event deadlines,
    (b) advance in sync in real time, and
    (c) can leave and join the network and remain symmetric:
%
\begin{enumerate}
\item \textbf{Event Deadlines:}
Due to the inherent latency of networks, a peer (through the middleware)
generates events with a deliberate $\Delta$ so that they can reach other peers
in time to be applied in sync.
%We first assume a predetermined delay for the sake of simplicity.
Nevertheless, a peer ahead of time may receive an event that should have been
applied in the past, which characterizes an optimistic approach to choose
$\Delta$, since it assumes that inconsistencies may occur~\cite{melding}.
%, even considering its $\Delta$.
%
In this case, the middleware automatically rolls back the peer, applies the
event at the correct time, and then fast forwards the peer, re-applying all
remaining events up to the real time.
As we detail further, rolling back a peer requires to simulate the whole
application execution and event occurrences from the beginning.
As an optimization, the middleware takes periodic snapshots locally to avoid
full simulation.
%
\item \textbf{Time Synchronization:}
Since peers run in different machines, their timelines will inevitably be out
of sync because applications are launched locally at different times, and
because internal clocks may diverge over time (e.g., from rollbacks or timer
polling inaccuracies).
The middleware assumes the maximum time among all peers the be the
``correct real time'', so that other peers must fast forward to catch up with
it.
Therefore, in order to determine the maximum time and synchronize the clocks,
the middleware proceeds as follows:
    (a) peers broadcast \code{SYNC} events periodically with their local
        times, and
    (b) if a peer is behind a received \code{SYNC}, it advances its frames
        proportionally faster to catch up in time.
%
To avoid abrupt visual changes, the middleware makes smooth time transitions
that take up to $1s$ to complete.
%
\item \textbf{Peer Churn:}
Considering that we target dynamic peer-to-peer networks, it is important that
the middleware can handle arrival and departure of peers seamlessly.
Note that high churn may lead to intermittent network partitions.
Given our unstructured approach, nothing needs to be done when a peer leaves
the network.
However, when a peer joins the network, the middleware needs to ensure that it
receives from other peer all events that ever happened, in order.
This is trivial since all peers already need to hold the full event history (as
described in item (a)).
% including their absolute timestamps in the shared timeline.
\end{enumerate}

As a limitation of our work, regarding items (a) and (b), we assume that peers
never lie about their local clocks, which would affect the event timestamps,
allowing a peer to modify the global timeline with nonsensical information.
%
For this reason, we assume a permissioned peer-to-peer network with an external
mechanism for authentication.

@@

Note that using callbacks is a already common technique for event-driven
programming, such as in Games, GUIs, and interactive applications in general,
sharing similarities with \emph{MVC}, \emph{Publish-Subscribe}, and
\emph{Observer} patterns~\cite{meyer,nystrom}.
%
Furthermore, the default run-to-completion semantics of callbacks is key to
ensure the desired determinism, since unlike threads, they run sequentially and
atomically from start to end~\cite{events,threads}.

@@

If dynamic allocation is required, it is necessary to hold a finite memory
pool in \code{G} with a custom allocator.
This is a limitation of C, since languages with serialization and
meta-programming offer better mechanisms to take memory snapshots.
Another alternative is to provide a serialization callback, which we left out
for future work.

@@

The middleware calls \code{cb\_ini} twice: at the beginning and at the end of
the loop.
The callback should initialize (and finalize) the network topology, as well as
static immutable globals that can live outside the simulation memory, such as
the window and image textures of our example:

@@

The simulation callback is the most sensitive API of the proposed middleware,
since it is responsible for preserving the consistency of the global
application state \code{G}, even under multiple executions from scratch.
%
Since the code is written in C, which is unaware of the semantics of symmetric
applications, note that it is the responsibility of the programmer to ensure
that the state is completely reproducible.
%
As discussed in Section~\ref{sec.related}, the typical approach, which we
chose, is to rely on deterministic and stateless calls only.

@@

The restriction for stateless calls is the reason why the rendering callback
\code{cb\_ren}~\cite{tml.js} must be separated from \code{cb\_sim}, which would
otherwise render the screen multiple times while time travelling.
Instead, the middleware calls \code{cb\_ren} to render the screen after each
frame event, but only in real time.
%, which the middleware calls only on real-time frames.
In our example, \code{cb\_ren} just needs to redraw the rectangle at the
current position:

@@

The callback \code{cb\_evt} is invoked by the middleware in only real time,
whenever a local SDL event occurs.
This callback has two goals:
    (a) to map and broadcast application events, and
    (b) to provide instant feedback to the user.
%As mentioned previously,
Not all local low-level events need to be broadcast in the network, only those
mapped as application events, as described in Section~\ref{sec.tml.api.events}.
The callback is allowed to perform side effects, such as modifying the network
topology, or exhibiting instant feedback on the screen.
In our example, we only generate application events for the 5 key of interest,
but we also exhibit the pressed key in real time as a visual
feedback:

@@

We now detail how the middleware orchestrates an application to ensure that all
peers
    (a) meet event deadlines,
    (b) advance in sync in real time, and
    (c) can leave and join the network and remain symmetric.

@@

As illustrated in Figure~\ref{fig.middleware}, events are broadcast between
peers with a deadline $\Delta$ scheduled to the future, such that all peers are
able to apply them in sync.
%
To prevent broadcast cycles, each peer keeps a set of the highest timestamps it
received from each other peer, such that lower numbers can be ignored when
received.
Since we rely on TCP connections, it is not possible to receive out-of-order
timestamps attached to a given source peer.
When the middleware receives a higher number from a given peer, it updates this
number and triggers a broadcast to neighbours.
%
%As mentioned in Section~\ref{sec.tml.api}, we assume that peers assign
%themselves unique identifiers in a contiguous range, such that they fit in a
%simple vector of timestamps.
%Therefore, this is not a severe limitation, and a separate mechanism could be
%used for assignments, with few modifications to the core of the middleware.

@@

The middleware main loop \code{p2p\_loop} is responsible for invoking the
application callbacks, and also controlling its timeline.
The loop continuously checks for network packets, local inputs, and time ticks,
as follows:

@@

Two facts ensure consistency in the network:
    (a) peers share a global timeline of events, and
    (b) they go through bit-identical steps
%
However, since the clocks may diverge temporarily, we must rely on simulation
rollbacks.
Therefore, the key to preserve consistency in the presence of rollbacks is to
restrict our programming API to deterministic and stateless calls only.
This way, it becomes impossible to diverge from other peers by, for example,
re-executing stateful file operations multiple times (or at different times).

@@

We run the application main loop at 50 FPS ($20ms$ per frame) and calculate
the middleware overhead at each iteration.
We found a negligible overhead of $9us$ in the average, which corresponds to
less than $1/2000$ of the CPU time.
This means that an application using the middleware would perform all network
synchronization and still have almost 100\% of the CPU time available for its
actual domain logic.
%
In order to have an absolute clock, we use a single machine to simulate all
peers, allowing us to merge their logs into a single file with a comparable
shared timeline.
%
We use \emph{NetEm}~\cite{netem} to simulate the network latency with a normal
distribution.
% (e.g., \code{tc qdisc add dev lo root netem delay 200ms 40ms distribution normal}).
%
We vary
    (i)   the network latency ($5-500ms$ between each peer),
    (ii)  the rate of events ($5-200~evt/sec$), and
    (iii) the deadline $\Delta$ multiplier ($1-5x$).
%
Figure~\ref{fig.mult} shows the $\Delta$s considering the network latency and
chosen multiplier.
For instance, a $50ms$ latency with a $2x$ multiplier results in
$\Delta=500ms$, i.e., each event is triggered with this deadline delay to
compensate the given latency and fixed average of 5 hops
($5 \times 50 \times 2 = 500$).
